---
title: "busclaR"
subtitle: "Bus lane priorities using GTFS and overline methods"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a method to support the prioritizartion of bus lanes.

This script helps to identify the road network segments with higher bus frequency, to support planning of BUS lanes.

It uses GTFS data for a given city or region, *overlines* the routes with they hourly frequency, and compares them with existing bus lanes and/or streets with more than 1 lane per direction.

We use mainly two R packages:

-   [**`GTFShift`**](https://u-shift.github.io/GTFShift/) to download and manipulate GTFS data
-   [**`stplanr`**](https://docs.ropensci.org/stplanr) to manipulate the bus routes

And the data:

-   GTFS for a given city or region, filtered by the modes that run on-street

-   OpenStreetMap for the same city or region, to compare with the existing Bus lanes.

In this paper we will use the Lisbon municipality as a case study.

## Methods

### Get GTFS data

Download and merge the existing GTFS for the Lisbon city

```{r}
# Lisbon_limit
aml = sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE)
lisboa = aml |> dplyr::filter(Concelho == "Lisboa") |> sf::st_bbox()

# get all GTFS
query_mobilitydatabase(token = Sys.getenv("MOBILITY_DATABASE"),
                       is_official = FALSE, # some feeds are not uploaded to the database by official sources
                       dataset_latitudes = c(lisboa[[2]], lisboa[[4]]),
                       dataset_longitudes = c(lisboa[[1]], lisboa[[3]]))
```
