---
title: "osm_bus_extract"
output: html_document
---

# Map shapes to OSM routes

```{r}
library(osmdata)
library(sf)
library(mapview)
library(dplyr)
library(GTFShift)

mapview::mapviewOptions(fgb = FALSE, viewer.suppress = TRUE)
```


## 1. Download OSM routes

```{r}
aml = sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE)
lisboa = aml |> dplyr::filter(Concelho == "Lisboa") |> sf::st_bbox()

q = opq(bbox = lisboa)  |> 
  add_osm_feature(key = "route", value = "bus") |> 
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)
```

Get routes multilines and stops
```{r}
carris_osm = q |> osmdata_sf()
carris_osm_multilines = carris_osm$osm_multilines
carris_osm_multilines_redux = carris_osm_multilines |> 
  select(osm_id, ref, from, to, via, name, roundtrip)
# mapview(carris_osm_multilines_redux) 

carris_osm_points = st_crop(carris_osm$osm_points, lisboa)
# mapview(carris_osm_points) 

carris_osm_stoppositions = carris_osm_points |> filter(public_transport == "stop_position") |> 
  select_if(~!all(is.na(.)))
# mapview(carris_osm_stoppositions, zcol = "ref") 

carris_osm_platform = carris_osm_points |> filter(public_transport == "platform") |> 
  select_if(~!all(is.na(.)))
# mapview(carris_osm_platform, zcol = "ref")
```

And relation between them

> Based on https://stackoverflow.com/questions/78758864/get-relation-data-from-osmdata

```{r}
# Save XML
library(xml2)
osm_file <- tempfile(fileext = ".osm")
osmdata_xml(q, filename = osm_file)

# Read and parse
doc <- read_xml(osm_file)

# Find all route relations
relations <- xml_find_all(doc, ".//relation[tag[@k='type'][@v='route'] and tag[@k='route'][@v='bus'] and tag[@k='network'][@v='Carris']]")

relations_df = lapply(relations, function(relation) {
  relation_id <- xml_attr(relation, "id")
  members <- xml_find_all(relation, "member")
  
  members_df = lapply(members, function(member) {
    c(  
       type = xml_attr(member, "type"),
       ref = xml_attr(member, "ref"),
       role = xml_attr(member, "role")
    )
  }) 
  df <- data.frame(do.call(rbind, members_df))
  df$relation_osm_id = relation_id
  return(df)
})
relations_df = bind_rows(relations_df)

relations_df |> distinct(role , .keep_all=TRUE)
```


## 2. Load GTFS
```{r}
data_sources = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))

gtfs_carris = tidytransit::read_gtfs(data_sources[data_sources$ID=="lisboa", ]$URL)
next_wednesday = calendar_nextBusinessWednesday(country_code="PT")
gtfs_carris = tidytransit::filter_feed_by_date(gtfs_carris, extract_date = next_wednesday)
# gtfs_carris = gtfs_carris |> filter_by_modes(modes = list(0, 3))  # filter by mode = tram and bus
```

## 3. Analyse one line

```{r}
carreira = "55B"
unique(carris_osm_multilines_redux$ref)
length(unique(carris_osm_multilines_redux$ref))
```

Filter osm network
```{r}
carris_osm_carreira = carris_osm_multilines_redux %>%
  filter(ref == carreira)

nrow(carris_osm_carreira) # 6
mapview(carris_osm_carreira, zcol="osm_id")
```

Filter GTFS
```{r}
shapes_sf = tidytransit::shapes_as_sf(gtfs_carris$shapes)
carris_gtfs_carreira = gtfs_carris$routes |>  # Start on routes.txt to match line number with route_short_name
  select(route_id, route_short_name, route_long_name) |>
  filter(route_short_name == carreira) |>
  left_join(gtfs_carris$trips |> select(route_id, trip_id, shape_id, direction_id), by="route_id") |> 
  left_join(shapes_sf, by="shape_id") |>
  distinct(shape_id, .keep_all = TRUE) |> 
  sf::st_as_sf()

View(carris_gtfs_carreira)
mapview(carris_gtfs_carreira)
nrow(carris_gtfs_carreira) # 6
```

## 4. Match based on initial and final points 

Compute osm final and initial points
```{r}
carris_osm_carreira_geom = carris_osm_carreira 

mapview(carris_osm_carreira_geom, zcol="osm_id")
st_geometry_type(carris_osm_carreira_geom)

carris_osm_carreira_geom = carris_osm_carreira_geom |> 
  rowwise() |>
  mutate(
    # Geographical data
    route_dist = st_length(geometry) |> units::drop_units(),
    # Other relevant parameters
    nr_stops = nrow(relations_df |> filter(relation_osm_id==osm_id & role=="stop")) + 2, # +2 because 1st and last stop have other role
    first_stop_osm_id = relations_df |> 
      # Consider both stop_entry/exit_only and stop, because circular lines do not have entry/exit, only stop
      filter(relation_osm_id==osm_id & role %in% c("stop_entry_only", "stop")) |> 
      # Use sorting to give priority to entry/exit, when they exist
      arrange(
        match(role, c("stop_entry_only", "stop")), 
        role
      ) |> 
      slice(1)  |> 
      pull(ref),
    last_stop_osm_id = relations_df |> 
      filter(relation_osm_id==osm_id & role %in% c("stop_exit_only", "stop")) |> 
      arrange(
        match(role, c("stop_exit_only", "stop")),
        role
      ) |> 
      slice(1) |> 
      pull(ref),
    initial = carris_osm_stoppositions |> filter(osm_id==first_stop_osm_id) |> slice(1) |> pull(geometry),
    final = carris_osm_stoppositions |> filter(osm_id==last_stop_osm_id) |> slice(1) |> pull(geometry)
  ) |> 
  ungroup() |> 
  select(osm_id, name, initial, final, route_dist, nr_stops, first_stop_osm_id, last_stop_osm_id, geometry) |> 
  arrange(route_dist, initial, final)

nrow(carris_osm_carreira_geom)
mapview(carris_osm_carreira_geom, zcol="osm_id")
carris_osm_carreira_geom_id = carris_osm_carreira_geom |> filter(osm_id == 19136075)
carris_osm_carreira_geom_id = carris_osm_carreira_geom |> filter(osm_id %in% c(8754460, 8754490))
mapview(carris_osm_carreira_geom_id, color="gray") + mapview(carris_osm_carreira_geom_id$initial, col.regions="#ffd400") + mapview(carris_osm_carreira_geom_id$final, col.regions="#003f8f")

mapview(carris_osm_carreira_geom, color="gray") + mapview(carris_osm_carreira_geom$initial, col.regions="#ffd400") + mapview(carris_osm_carreira_geom$final, col.regions="#003f8f")
```

Same for GTFS shapes
```{r}
# 2. Match gtfs shapes and osm routes, by choosing the one that share the closest start and end points
carris_gtfs_carreira = carris_gtfs_carreira |> 
  rowwise() |>
  mutate(
    # Geographical data
    route_dist = st_length(geometry) |> units::drop_units(),
    initial = lwgeom::st_startpoint(geometry), # We could also replace this with stop.txt locations
    final = lwgeom::st_endpoint(geometry),
    # Other relevant parameters
    trip_id_copy = trip_id,
    nr_stops = nrow(gtfs_carris$stop_times |> filter(trip_id == trip_id_copy))
  ) |> 
  ungroup() |> 
  select(-trip_id_copy) |>
  arrange(route_dist, initial, final)

carris_gtfs_carreira_id = carris_gtfs_carreira |> filter(grepl("171_3", shape_id))
mapview(carris_gtfs_carreira_id, color="gray") + mapview(carris_gtfs_carreira_id$initial, col.regions="#ffd400") + mapview(carris_gtfs_carreira_id$final, col.regions="#003f8f")

mapview(carris_gtfs_carreira, color="gray") + mapview(carris_gtfs_carreira$initial, col.regions="#ffd400") + mapview(carris_gtfs_carreira$final, col.regions="#003f8f")

```

Compute distances between init and final points for both
```{r}
init = st_distance(carris_osm_carreira_geom$initial, carris_gtfs_carreira$initial)
fin = st_distance(carris_osm_carreira_geom$final, carris_gtfs_carreira$final)
```

Match OSM network and GTFS shapes considering the match with min aggregated distance (init + fin)
```{r}
conjunto = abs(init+fin)
conjunto # matrix osm (lines) x gtfs (cols)

carris_gtfs_carreira_minimos = carris_gtfs_carreira |> 
  mutate(osm_id = NA)

for (i in 1:nrow(carris_gtfs_carreira_minimos)) {
  carris_gtfs_carreira_minimos[i,]$osm_id = carris_osm_carreira_geom[which.min(conjunto[,i]),]$osm_id
}

carris_gtfs_carreira_result = carris_gtfs_carreira_minimos |>
  st_drop_geometry() |>  
  left_join(
    carris_osm_carreira_geom |> select(osm_id, name, route_dist, nr_stops, geometry), 
    by = "osm_id",
    suffix = c("_gtfs", "_osm")
  ) |> 
  mutate(distance_diff = abs(route_dist_gtfs - route_dist_osm)) |> # absolute difference
  st_as_sf(sf_column_name="geometry")


carris_gtfs_carreira_result$shape_osm = paste(
  carris_gtfs_carreira_result$shape_id, 
  " // ", 
  carris_gtfs_carreira_result$osm_id , 
  " // ", 
  carris_gtfs_carreira_result$distance_diff
)
```

Analyse results
```{r}
View(carris_gtfs_carreira_result)
max(carris_gtfs_carreira_result$distance_diff)

mapview(carris_gtfs_carreira, zcol = "shape_id")
mapview(carris_gtfs_carreira_result, zcol = "shape_osm")

mapview(carris_gtfs_carreira |> filter(grepl("106_4", shape_id)), zcol = "shape_id")
mapview(carris_gtfs_carreira_result |> filter(grepl("106_4", shape_id)), zcol = "shape_osm")
```

```{r}
names(carris_gtfs_carreira_result)

carris_gtfs_carreira_result |>
  select(shape_id, nr_stops_gtfs, route_dist_gtfs, osm_id, nr_stops_osm, route_dist_osm ,distance_diff)
```

 